\documentclass{homework}
\usepackage{fullpage}
\usepackage[super]{nth}



\title{MATH 4350 -- Final Exam}
\author{Stuart Olsen}
\date{May 11, 2015}

\begin{document}
\maketitle



\section*{Part I}
\problem{1}{Let $S = \{1, 2, 3, 4, 5, 6\}$.}
\subproblem{a}{How many $3$-permutations of $S$ are there?}
The number of $3$-permutations of a $6$-set is given by
\begin{align*}
  P(6, 3) &= 6 \cdot 5 \cdot 4\\
          &= \answer{120}.
\end{align*}

\subproblem{b}{How many $3$-combinations of $S$ are there?}
The number of $3$-combinations of a $6$-set is given by
\begin{align*}
  {6 \choose 3} &= \frac{P(6, 3)}{3!}\\
                &= \frac{120}{6}\\
                &= \answer{20}.
\end{align*}

\subproblem{c}{How many cyclic $3$-permutations of $S$ are there?}
The number of cyclic $3$-permutations of a $6$-set is given by
\begin{align*}
  \frac{P(6, 3)}{3} &= \frac{120}{3}\\
                    &= \answer{40}.
\end{align*}



\problem{2}{Let $S = \{a, a, b, b, b, c\}$.}
\subproblem{a}{How many $3$-permutations of $S$ are there?}
Let $f(r,S)$ denote the number of $r$-permutations of the multiset $S$. Then,
\begin{align*}
  f(3,S) &= f(3,\{2 \cdot a, 3 \cdot b, 1 \cdot c\})\\
         &= f(2,\{1 \cdot a, 3 \cdot b, 1 \cdot c\})\\
         &+ f(2,\{2 \cdot a, 2 \cdot b, 1 \cdot c\})\\
         &+ f(2,\{2 \cdot a, 3 \cdot b\})\\
         &= f(1,\{3 \cdot b, 1 \cdot c\}) + f(1,\{1 \cdot a, 2 \cdot b, 1 \cdot c\}) + f(1,\{1 \cdot a, 3 \cdot b\})\\
         &+ f(1,\{1 \cdot a, 2 \cdot b, 1 \cdot c\}) + f(1,\{2 \cdot a, 1 \cdot b, 1 \cdot c\}) + f(1,\{2 \cdot a, 2 \cdot b\})\\
         &+ f(1,\{1 \cdot a, 3 \cdot b\}) + f(1,\{2 \cdot a, 2 \cdot b\})\\
         &= (2 + 3 + 2) + (3 + 3 + 2) + (2 + 2)\\
         &= 7 + 8 + 4\\
         &= \answer{19}.
\end{align*}

\subproblem{b}{How many $3$-combinations of $S$ are there?}
\incomplete{
  Let $f(r,S)$ denote the number of $r$-combinations of the multiset $S$. Then,
  \begin{align*}
    f(3,S) &= f(3,\{2 \cdot a, 3 \cdot b, 1 \cdot c\})
  \end{align*}
}



\problem{3}{
  Use the pigeonhole principle to show that if there are $13$ students in a class then at least $4$ students get the same grade (assume the only grades assigned are A, B, C, D, or F).
}
This is not the case; it is possible, for example, for $3$ students to receive an A, $3$ students to receive a B, $3$ students to receive a C, $3$ students to receive a D, and one student to receive an F.

If, however, the only grades assigned are A, B, C, and F, we can interpret grades as boxes and students' individual grades as objects to be placed in those boxes. Then, by the strong form of the Pigeonhole Principle, in distributing $4 + 4 + 4 + 4 - 4 + 1 = 13$ objects into four boxes (that is, assigning one of four grades to each of $13$ students), we are guaranteed that at least four objects (students) are assigned to some box (grade).



\problem{4}{State the Binomial Theorem and use it to expand $(x+y)^3$.}
The Binomial Theorem states that for all positive integers $n$ and all $x$ and $y$,
\begin{align*}
  (x+y)^n = \sum_{k=0}^n{{n \choose k}x^{n-k}y^k}.
\end{align*}
Then,
\begin{align*}
  (x+y)^3 &= \sum_{k=0}^3{{3 \choose k}x^{3-k}y^k}\\
  &= {3 \choose 0}x^{3-0}y^0 + {3 \choose 1}x^{3-1}y^1 + {3 \choose 2}x^{3-2}y^2 + {3 \choose 3}x^{3-3}y^3\\
  &= \frac{3!}{3!\,0!}x^3 + \frac{3!}{2!\,1!}x^2y + \frac{3!}{1!\,2!}xy^2 + \frac{3!}{0!\,3!}y^3\\
  &= \answer{x^3 + 3x^2y + 3xy^2 + y^3}.
\end{align*}



\problem{5}{Evaluate ${6 \choose 3\ 2\ 1}$.}
\begin{align*}
  {6 \choose 3\ 2\ 1} &= \frac{6!}{3!\,2!\,1!}\\
  &= \frac{6 \cdot 5 \cdot 4 \cdot 3 \cdot 2 \cdot 1}{(3 \cdot 2 \cdot 1) \cdot (2 \cdot 1) \cdot (1)}\\
  &= \frac{720}{6 \cdot 2 \cdot 1}\\
  &= \answer{60}.
\end{align*}



\problem{6}{
  $100$ people are surveyed and asked if they are planning on purchasing a laptop or a PC during the next year. Of the people surveyed, $40$ people are planning on buying a laptop and $30$ are planning on buying a PC, while $10$ are planning on buying both. State the Principle of Inclusion-Exclusion for $2$ sets and use it to determine how many people are planning on buying neither a laptop nor a PC.
}
The Inclusion-Exclusion Principle for two subsets $A_1$ and $A_2$ of $S$ is
\begin{align*}
  \card{\compl{A_1} \cap \compl{A_2}} &= \card{S} - (\card{A_1} + \card{A_2}) + \card{A_1 \cap A_2}.
\end{align*}
Then, if we let $S$ be the set of people surveyed, $A_1$ the set of those planning on buying a laptop, and $A_2$ the set of those planning on buying a PC, the number of people planning on buying neither a laptop nor a PC is given by
\begin{align*}
  \card{\compl{A_1} \cap \compl{A_2}} &= \card{S} - (\card{A_1} + \card{A_2}) + \card{A_1 \cap A_2}\\
  &= 100 - (40 + 30) + 10\\
  &= \answer{40}.
\end{align*}



\problem{7}{
  Determine the generating function and the exponential generating function for the sequence $\{1, 1, 1, 1, \ldots\}$. Make sure to label both.
}
The generating function for the above sequence is
\begin{align*}
  g(x) &= 1 + x + x^2 + x^3 + \cdots\\
  &= \answer{\frac{1}{1-x}}.
\end{align*}
The exponential generating function is
\begin{align*}
  g^{(e)}(x) &= \sum_{n=0}^\infty{\frac{x^n}{n!}}\\
  &= \answer{e^x}.
\end{align*}



\problem{8}{Answer the following for the sequence $h_n = n^2$.}
\subproblem{a}{Find the difference sequence of $h_n$ using $n \in \{0, 1, 2, 3 ,4\}$.}
We have
\begin{align*}
  \Delta h_n &= h_{n+1} - h_n\\
  &= (n+1)^2 - n^2\\
  &= (n^2 + 2n + 1) - n^2\\
  &= 2n + 1.
\end{align*}
Then,
\begin{align*}
  \Delta h = \answer{\{1, 3, 5, 7, 9, \ldots\}}.
\end{align*}

\subproblem{b}{Using the $\nth{0}$ diagonal, determine the Stirling numbers of the second kind $S(2,0)$, $S(2,1)$, and $S(2,2)$.}
We have
\begin{align*}
  h &= \{0, 1, 4, 9, 16, 25, \dots\};\\
  \Delta h &= \{1, 3, 5, 7, 9, \ldots\};\\
  \Delta^2 h &= \{2, 2, 2, 2, \ldots\};
\end{align*}
so
\begin{align*}
  S(2,0) &= \frac{0}{0!} = \answer{0};\\
  S(2,1) &= \frac{1}{1!} = \answer{1};\\
  S(2,2) &= \frac{2}{2!} = \answer{1}.
\end{align*}



\problem{9}{Find the Stirling numbers of the first kind $s(2,0)$, $s(2,1)$, and $s(2,2)$ by expanding $\fall{n}{2}$.}
We are given:
\begin{align*}
  \fall{n}{p} = \sum_{k=0}^p{(-1)^{p-k}s(p,k)n^k}.
\end{align*}
Then,
\begin{align*}
  \fall{n}{p} &= n(n-1)\\
  &= n^2 - n,
\end{align*}
so
\begin{align*}
  (-1)^2s(2,0)n^0 &= 0;\\
  (-1)^1s(2,1)n^1 &= -n;\\
  (-1)^0s(2,2)n^2 &= n^2
\end{align*}
or
\begin{align*}
  s(2,0) &= \answer{0};\\
  s(2,1) &= \answer{1};\\
  s(2,2) &= \answer{1}.
\end{align*}



\problem{10}{
  Let $\mathcal{A} = \{A_1, A_2, A_3, A_4\}$, where
  \begin{align*}
    A_1 &= \{a, c, d\};\\
    A_2 &= \{a, b, c\};\\
    A_3 &= \{a, d\};\\
    A_4 &= \{a, b, d\}.
  \end{align*}
  List an SDR for $\mathcal{A}$.
}
$\answer{\{c, b, d, a\}}$.



\problem{11}{
  Create an addition and multiplication table for $\bz_4$. List the identities for each operation. Explain which operation is a Latin square. Explain which operation does not have an inverse.
}
\begin{center}
  \begin{tabular}{c|cccc}
    ${\oplus}$ & $0$ & $1$ & $2$ & $3$\\
    \hline
           $0$ & $0$ & $1$ & $2$ & $3$\\
           $1$ & $1$ & $2$ & $3$ & $0$\\
           $2$ & $2$ & $3$ & $0$ & $1$\\
           $3$ & $3$ & $0$ & $1$ & $2$\\
  \end{tabular}\quad\quad
  \begin{tabular}{c|cccc}
    ${\otimes}$ & $0$ & $1$ & $2$ & $3$\\
    \hline
            $0$ & $0$ & $0$ & $0$ & $0$\\
            $1$ & $0$ & $1$ & $2$ & $3$\\
            $2$ & $0$ & $2$ & $0$ & $2$\\
            $3$ & $0$ & $3$ & $2$ & $1$\\
  \end{tabular}
\end{center}

The identity of ${\oplus}$ is $\answer{0}$ and that of ${\otimes}$ is $\answer{1}$. ${\oplus}$ yields a Latin square, but ${\otimes}$ does not as $1$ does not appear in row $0$. ${\otimes}$ does not have an inverse, as there is no $x \in \bz_4$ such that $2 \otimes x = 1$.



\problem{12}{Is there a BIBD with parameters $b=6$, $k=6$, $v=11$, $r=4$, and some $\lambda$?}
Suppose there is some such BIBD. Then,
\begin{align*}
  bk &= vr\\
  6 \cdot 6 &= 11 \cdot 4\\
  36 &= 44,
\end{align*}
which is false. Therefore, there is no such BIBD.



\section*{Part II}
\problem{1}{
  Using the Principle of Inclusion-Exclusion, find the number of integral solutions to
  \begin{align*}
    x_1 + x_2 + x_3 = 3
  \end{align*}
  given $1 \le x_1 \le 3$, $0 \le x_2 \le 1$, and $0 \le x_3 \le 1$.
}
Let
\begin{align*}
  y_1 &= x_1 - 1;\\
  y_2 &= x_2;\\
  y_3 &= x_3.
\end{align*}
Then, the given problem is equivalent to finding the number of integral solutions to
\begin{align}
  y_1 + y_2 + y_3 = 2 \label{eq:II1y}
\end{align}
where
\begin{align*}
  0 \le\ &y_1 \le 2;\\
  0 \le\ &y_2 \le 1;\\
  0 \le\ &y_3 \le 1.
\end{align*}

Let $S$ be the set of non-negative integral solutions to Equation~\eqref{eq:II1y}. Then, $\card{S} = {2 + 3 - 1 \choose 2} = {4 \choose 2} = 6$. For each $i$ in $\{1,2,3\}$, let $A_i \subseteq S$ be the set of solutions wherein the above constraint on $y_i$ does not hold. Then, the solution to the given problem is given by
\begin{align*}
  \card{\compl{A_1} \cap \compl{A_2} \cap \compl{A_3}} &= \card{S}\\
                                                       &- (\card{A_1} + \card{A_2} + \card{A_3})\\
                                                       &+ (\card{A_1 \cap A_2} + \card{A_1 \cap A_3} + \card{A_2 \cap A_3})\\
&- (\card{A_1 \cap A_2 \cap A_3}).
\end{align*}
Note that $\card{A_1} = 0$. Then,
\begin{align*}
  \card{\compl{A_1} \cap \compl{A_2} \cap \compl{A_3}} &= \card{S} - (\card{A_2} + \card{A_3}) + \card{A_2 \cap A_3}.
\end{align*}
Because there are only $2$ units to distribute among $y_1$, $y_2$, and $y_3$, there is only one solution in which $y_2 \nleq 1$: that in which $y_1 = 0$, $y_2 = 2$, and $y_3 = 0$. Therefore, $\card{A_2} = 1$, and by the same reasoning, $\card{A_3} = 1$. Similarly, to satisfy $y_2 \nleq 1$ and $y_3 \nleq 1$ would require at least $4$ units to distribute. Therefore, $\card{A_2 \cap A_3} = 0$. Thus,
\begin{align*}
  \card{\compl{A_1} \cap \compl{A_2} \cap \compl{A_3}} &= 6 - (1 + 1) + 0\\
                                                       &= \answer{4}.
\end{align*}



\problem{2}{
  You wish to make a fruit basket consisting of apples, oranges, and bananas. Find the generating function whose coefficients determine the number of fruit baskets with $n$ pieces of fruit given that there must be between $1$ and $3$ apples, at most $1$ orange, and at most $1$ banana. Use your result to determine the number of fruit baskets containing exactly $3$ pieces of fruit.
}
We have
\begin{align*}
  g_{apple}(x) &= x + x^2 + x^3\\
  g_{orange}(x) &= 1 + x\\
  g_{banana}(x) &= 1 + x.
\end{align*}
Combining yields
\begin{align*}
  g(x) &= (x + x^2 + x^3)(1 + x)(1 + x)\\
       &= (x + x^2 + x^3)(1 + 2x + x^2)\\
       &= x + 2x^2 + x^3 + x^2 + 2x^3 + x^4 + x^3 + 2x^4 + x^5\\
       &= x + 3x^2 + 4x^3 + 3x^4 + x^5.
\end{align*}
Thus, there are $\answer{4}$ ways to make a basket containing exactly $3$ pieces of fruit.



\problem{3}{
  Using the method indicated, count the number of ways to color an $n \times 1$ board with colors RED and BLUE such that at least one square must be BLUE.
}
\subproblem{a}{By basic counting principles and counting the complement.}
Because there are $n$ squares and each square may receive one of two colors, there are $2^n$ ways to color the board without restrictions. The complement of the given set of restrictions is simply that no square is BLUE, i.e. that every square is RED; there is one way of meeting this restriction. Thus, there are $\answer{2^n - 1}$ ways to color an $n \times 1$ board with colors RED and BLUE such that at least one square is BLUE.

\subproblem{b}{
  By finding the exponential generating function for $h_n$, which is equal to the number of ways to color an $n \times 1$ board with colors RED and BLUE such that at least one square must be BLUE.
}
Because there may be any number of RED squares and any number of blue squares greater than zero, we have
\begin{align*}
  {g_{\mathrm{RED}}}^{(e)}(x) &= 1 + \frac{x}{1!} + \frac{x^2}{2!} + \cdots = e^x\\
  {g_{\mathrm{BLUE}}}^{(e)}(x) &= \frac{x}{1!} + \frac{x^2}{2!} + \frac{x^3}{3!} + \cdots\\
                              &= (1 + \frac{x}{1!} + \frac{x^2}{2!} + \cdots) - 1\\
                              &= e^x - 1.
\end{align*}
Then,
\begin{align*}
  g^{(e)}(x) &= e^x(e^x - 1)\\
             &= (e^x)^2 - e^x\\
             &= e^{2x} - e^x\\
             &= \sum_{n=0}^\infty{\frac{(2x)^n}{n!}} - \sum_{n=0}^\infty{\frac{x^n}{n!}}\\
             &= \sum_{n=0}^\infty{2^n\frac{x^n}{n!}} - \sum_{n=0}^\infty{\frac{x^n}{n!}}\\
             &= \sum_{n=0}^\infty{(2^n-1)\frac{x^n}{n!}}.
\end{align*}
Therefore, $h_n = \answer{2^n-1}$.



\problem{4}{Find a closed form for $h_n = 3h_{n-1} - 2h_{n-2}$ given the initial values $h_0 = 0$ and $h_1 = 1$ using the indicated method.}
\subproblem{a}{Suppose $h_n = q^n$.}
We have
\begin{align*}
  q^n &= 3q^{n-1} - 2q^{n-2}\\
  (q^2)q^{n-2} &= (3q)q^{n-2} - 2q^{n-2}\\
  q^2 &= 3q - 2\\
  q^2 - 3q + 2 &= 0.
\end{align*}
Then,
\begin{align*}
  q &= \frac{3 \pm \sqrt{9 - 8}}{2}\\
    &= \frac{3 \pm 1}{2}\\
    &= 1, 2.
\end{align*}
Then, there are some $c_0$ and $c_1$ such that
\begin{align*}
  h_n &= c_0 2^n + c_1 1^n\\
      &= c_0 2^n + c_1.
\end{align*}
Then, solving the system of linear equations
\begin{align*}
  \begin{cases}
    (n=0) & c_0 + c_1 = 0\\
    (n=1) & 2c_0 + c_1 = 1
  \end{cases}
\end{align*}
yields
\begin{align*}
  c_0 &= 1;\\
  c_1 &= -1.
\end{align*}
Thus,
\begin{align*}
  \answer{h_n = 2^n - 1}.
\end{align*}

\subproblem{b}{Use generating function $g(x) = h_0 + h_1x + h_2x^2 + \cdots$ and sum multiples of $g$ to produce the homogeneous recursion.}
We have
\begin{align*}
  h_n - 3h_{n-1} + 2h_{n-2} = 0.
\end{align*}
Then, noting
\begin{align*}
  g(x) &= \sum_{n=0}^\infty{h_nx^n}\\
       &= h_0 + h_1x + \sum_{n=2}^\infty{h_nx^n};
\end{align*}
\begin{align*}
  -3xg(x) &= -3x\sum_{n=0}^\infty{h_nx^n}\\
       &= \sum_{n=0}^\infty{-3h_nx^{n+1}}\\
       &= -3h_0x + \sum_{n=1}^\infty{-3h_nx^{n+1}}\\
       &= -3h_0x + \sum_{n=2}^\infty{-3h_{n-1}x^n};
\end{align*}
\begin{align*}
  2x^2g(x) &= 2x^2\sum_{n=0}^\infty{h_nx^n}\\
       &= \sum_{n=0}^\infty{2h_nx^{n+2}}\\
       &= \sum_{n=2}^\infty{2h_{n-2}x^n},
\end{align*}
we have
\begin{align*}
  g(x) - 3xg(x) + 2x^2g(x) &= h_0 + h_1x - 3h_0x + \sum_{n=2}^\infty{(h_n - 3h_{n-1} + 2h_{n-2})x^n}\\
  (1 - 3x + 2x^2)g(x) &= h_0 + h_1x - 3h_0x + \sum_{n=2}^\infty{(0)x^n}\\
                           &= 0 + x - 0 + 0\\
                           &= x,
\end{align*}
so
\begin{align*}
  g(x) &= \frac{x}{1 - 3x + 2x^2}\\
       &= \frac{x}{(2x-1)(x-1)}.
\end{align*}
Proceeding with partial fraction decomposition:
\begin{align*}
  g(x) &= \frac{x}{(2x-1)(x-1)}\\
       &= \frac{A}{2x-1} + \frac{B}{x-1}\\
  x &= A(x-1) + B(2x-1)\\
       &= Ax - A + 2Bx - B\\
       &= (A + 2B)x - (A + B)\\
  0 &= (A + 2B - 1)x - (A + B)\\
  A + B &= (A + 2B - 1)x
\end{align*}
yields the system of equations
\begin{align*}
  \begin{cases}
    A + B &= 0\\
    A + 2B - 1 &= 0,
  \end{cases}
\end{align*}
which has the solution
\begin{align*}
  A &= -1;\\
  B &= 1.
\end{align*}
Then,
\begin{align*}
  g(x) &= -\frac{1}{2x - 1} + \frac{1}{x - 1}\\
       &= \frac{1}{1 - 2x} - \frac{1}{1 - x}\\
       &= \sum_{n=0}^\infty{(2x)^n} - \sum_{n=0}^\infty{x^n}\\
       &= \sum_{n=0}^\infty{(2^n - 1)x^n}.
\end{align*}
Thus,
\begin{align*}
  \answer{h_n = 2^n - 1}.
\end{align*}



\problem{5}{
  Let $\mathcal{A} = \{A_1, A_2, A_3, A_4, A_5\}$, where
  \begin{align*}
    A_1 &= \{a, c, d\};\\
    A_2 &= \{a, c\};\\
    A_3 &= \{a, d\};\\
    A_4 &= \{a, b, d\};\\
    A_5 &= \{c, d\}.
  \end{align*}
  Determine whether $\mathcal{A}$ has an SDR using the Marriage Condition.
}
We have
\begin{align*}
  \card{A_1 \cup A_2 \cup A_3 \cup A_4} &= \card{\{a, c, d\} \cup \{a, c\} \cup \{a, d\} \cup \{c, d\}}\\
                                        &= \card{\{a, c, d\}}\\
                                        &= 3\\
                                        &\ngeq 4;
\end{align*}
therefore, the marriage condition does not hold for $\mathcal{A}$, and so $\mathcal{A}$ does {\em not} have an SDR.



\problem{6}{Which block, $B_1 = \{3, 4, 6\}$ or $B_2 = \{0, 3, 4\}$, generates a BIBD over $\bz_7$?}
\incomplete{}



\problem{7}{
  Answer the following for
  \begin{align*}
    L = \begin{bmatrix}
      0 & 1 & 2\\
      1 & 2 & 0\\
      2 & 0 & 1
    \end{bmatrix}.
  \end{align*}
}
\subproblem{a}{Verify $L$ is a Latin square using the matrices $R_3$ and $S_3$.}
\incomplete{}
\subproblem{b}{Show $L$ is in the $L_3^r$ family and determine $r$.}
\incomplete{}
\subproblem{c}{Find a Latin square orthogonal to $L$ and verify orthogonality.}
\incomplete{}



\problem{8}{
  Prove Pascal's Formula:
  \begin{align*}
    {n \choose k} = {n - 1 \choose k} + {n - 1 \choose k - 1}
  \end{align*}
  using the following proof methods.
}
\subproblem{a}{Algebraic proof.}
\incomplete{}
\subproblem{a}{Combinatorial proof.}
\incomplete{}



\problem{9}{Let $h_n = n^p$. Using induction, show that $\Delta^{p+1}(h_n) = 0$ for all $n$.}
\incomplete{}



\end{document}