\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amsfonts}

\title{MATH 4350 -- Homework 1}
\author{Stuart Olsen}

\begin{document}
\maketitle

\section*{Chapter 2.7}
\subsection*{1. For each of the four subsets of the two properties (a) and (b), count the number of four-digit numbers whose digits are either $1$, $2$, $3$, $4$, or $5$:}
\subsubsection*{(a) The digits are distinct.}
\subsubsection*{(b) The number is even.}

\begin{itemize}
\item Case $\{\}$:

There are five possibilities for each of four independent digits; the number of possibilities is then $5^4$.
\item Case $\{a\}$:

There are five possibilities for the first digit. For each, there are four possibilities for the second; for each of these, three possibilities for the third; and for each of these, two for the last. Then, the number of possibilities is $5 \cdot 4 \cdot 3 \cdot 2$.
\item Case $\{b\}$:

A number is even iff its last digit is even. Then, there are two possibilities ($2$ and $4$) for the fourth digit and five for each of the rest. This gives $5^3 \cdot 2$ possibilities.
\item Case $\{a, b\}$:

There are two possibilities for the last digit. For each, there are four possibilities for the third digit, for each of which there are three possibilities for the second digit, for each of which there are two possibilities for the third. Then, there are $2 \cdot 4 \cdot 3 \cdot 2$ possibilities.
\end{itemize}

\subsection*{4. How many distinct positive divisors does each of the following numbers have?}
\subsubsection*{(a) $3^4 \times 5^2 \times 7^6 \times 11$}
Each divisor can be written as $3^i \cdot 5^j \cdot 7^m \cdot 11^n$, for some $0 \le i \le 4$, $0 \le j \le 2$, $0 \le m \le 6$, and $0 \le n \le 1$. There are $5 \cdot 3 \cdot 7 \cdot 2$ such expressions. By the fundamental theorem of arithmetic, this is exactly the number of divisors.
\subsubsection*{(b) $620$}
The prime factorization of $620$ is $2^2 \times 5 \times 31$. By the above process, there are $3 \cdot 2 \cdot 2$ divisors.
\subsubsection*{(c) $10^{10}$}
The prime factorization of $10^{10}$ is $2^{10} \times 5^{10}$. By the above process, there are $11 \cdot 11$ divisors.

\subsection*{7. In how many ways can four men and eight women be seated at a round table if there are to be two women between consecutive men around the table?}
Choose some $m_1$ from the set of men. There are $12$ choices for seating $m_1$. For each, there are three seats in which the three remaining men may sit; then, for each seating for $m_1$, there are $3!$ ways to seat the men. For each way to seat the men, there are (from $8$ women and $8$ seats) $8!$ ways to seat the women. Thus, there are $12 \cdot 3! \cdot 8!$ ways to seat the party.

If seatings which differ only by rotation are considered equivalent, then there are $\frac{12 \cdot 3! \cdot 8!}{12} = 3! \cdot 8!$ ways to seat the party.

\subsection*{10. A committee of five people is to be chosen from a club that boasts a membership of $10$ men and $12$ women. How many ways can the committee be formed if it is to contain at least two women? How many ways if, in addition, one particular man and one particular woman who are members of the club refuse to serve together on the committee?}
There are ${12 + 10} \choose 5$ ways to form a committee of five with no restrictions on membership. Of these, ${12 \choose 0} \cdot {10 \choose 5}$ contain no women and ${12 \choose 1} \cdot {10 \choose 4}$ contain only one. Then, there are ${{12 + 10} \choose 5} - \left[ {{12 \choose 0} \cdot {10 \choose 5}} + {{12 \choose 1} \cdot {10 \choose 4}} \right]$ ways to form a committee of five with at least two women.

On a committee with $n$ women ($2 \le n \le 5$), there are ${12 - 1} \choose {n - 1}$ ways to choose the women such that the particular woman is on the committee; and ${10 - 1} \choose {5 - n - 1}$ ways to choose the men such that the particular man is on the committee. Since these are independent, there are ${11 \choose {n - 1}} \cdot {9 \choose {4 - n}}$ ways to form a committee with the particular man and woman. Thus, there are ${{12 + 10} \choose 5} - \left[ {{12 \choose 0} \cdot {10 \choose 5}} + {{12 \choose 1} \cdot {10 \choose 4}} \right] - \left[ \sum_{n=2}^5 {11 \choose {n - 1}} \cdot {9 \choose {4 - n}} \right]$ ways to form a committee meeting all restrictions.

\subsection*{11. How many sets of three integers between $1$ and $20$ are possible if no two consecutive integers are to be in a set?}
There are $20 \choose 3$ subsets of $\mathbb{Z} \cap [1, 20]$ of size $3$. These can be partitioned into those which contain no consecutive elements, those which contain a pair of consecutive elements, and those which contain three consecutive elements.

There are $17$ subsets containing $1$, $2$, and another non-consecutive element; the same holds for $19$ and $20$. For each of the remaining $17$ consecutive pairs, there are $16$ such subsets. There are $18$ subsets containing three consecutive elements. Thus, there are ${20 \choose 3} - \left[ 2 \cdot 17 + 17 \cdot 16 \right] - 18$ subsets of $\mathbb{Z} \cap [1, 20]$ of size three containing no consecutive elements.

\subsection*{15. At a party there are $15$ men and $20$ women.}
\subsubsection*{(a) How many ways are there to form $15$ couples consisting of one man and one woman?}
There are ${15 \choose 15} = 1$ ways to choose the men for the couples. For each, there are ${20 \choose 1} = 20$ ways to choose a woman to pair with the first man, ${19 \choose 1} = 19$ ways to choose a woman to pair with the second, and so on. Then, there are $1 \cdot (20 \cdot 19 \cdot \cdots \cdot 7 \cdot 6)$ ways to form $15$ couples.

\subsubsection*{(b) How many ways are there to form $10$ couples consisting of one man and one woman?}
There are ${15 \choose 10}$ ways to choose the men for the couples. For each, there are ${20 \choose 1} = 20$ ways to choose a woman to pair with the first man, ${19 \choose 1} = 19$ ways to choose a woman to pair with the second, and so on. Then, there are ${15 \choose 10} \cdot (20 \cdot 19 \cdot \cdots \cdot 12 \cdot 11)$ ways to form $10$ couples.

\subsection*{17. In how many ways can six indistinguishable rooks be placed on a $6$-by-$6$ board so that no two rooks can attack one another? In how many ways if there are two red and four blue rooks?}
No two rooks may be on the same row or column. Because there are six rooks, six rows, and six columns, there must be exactly one column associated with each row and vice versa. Then, there are $6$ possible columns to pair with the first row; for each, there are $5$ to pair with the second; and so on. Then, there are $6 \cdot 5 \cdot 4 \cdot 3 \cdot 2 \cdot 1$ arrangements.

Because color and position are independent, we can multiply the number of position arrangements by the number of color arrangements for a single position arrangement. Because there are ${6 \choose 2} = {6 \choose 4}$ ways to color a position arrangement, there are then $6! \cdot {6 \choose 2}$ position-color arrangements.

\subsection*{20. Determine the number of circular permutations of $\{0, 1, 2, \ldots, 9\}$ in which $0$ and $9$ are not opposite. ({\it Hint}: Count those in which $0$ and $9$ are opposite.)}
There are $10!$ (non-circular) permutations of $\{0, 1, 2, \ldots, 9\}$. For each of the $10$ positions in which $0$ can be placed, there are $8!$ possible permutations in which $0$ and $9$ are ``opposite.'' Then, there are $10! - 10 \cdot 8!$ permutations in which $0$ and $9$ are {\it not} opposite. Dividing this by $10$ yields $9! - 8!$ circular permutations in which $0$ and $9$ are not opposite.

\subsection*{21. How many permutations are there of the letters of the word ADDRESSES? How many $8$-permutations are there of these nine letters?}
With letter uniqueness, there are $9!$ permutations of the word ADDRESSES. Then, there are $\frac{9!}{1! \cdot 2! \cdot 1! \cdot 2! \cdot 3!}$ without.

The 8-permutations obtained by omitting one of each letter have the following numbers:
\begin{itemize}
\item A: $\frac{8!}{0! \cdot 2! \cdot 1! \cdot 2! \cdot 3!}$
\item D: $\frac{8!}{1! \cdot 1! \cdot 1! \cdot 2! \cdot 3!}$
\item R: $\frac{8!}{1! \cdot 2! \cdot 0! \cdot 2! \cdot 3!}$
\item E: $\frac{8!}{1! \cdot 2! \cdot 1! \cdot 1! \cdot 3!}$
\item S: $\frac{8!}{1! \cdot 2! \cdot 1! \cdot 2! \cdot 2!}$
\end{itemize}

The total number of 8-permutations is given by the sum of these.

\subsection*{24. A roller coaster has five cars, each containing four seats, two in front and two in back. There are $20$ people ready for a ride. In how many ways can the ride begin? What if a certain two people want to sit in different cars?}
Because people and seats are each distinguishable, the number of unrestricted configurations is given by $20!$ ($20$ people permuted across $20$ seats). For each of the five cars, there are $4 \cdot 3$ ways to seat the certain people together in that car. For each such seating, there are $18!$ ways to seat the remaining passengers. Then, there are $5 \cdot 4 \cdot 3 \cdot 18!$ ways to put the certain two people in the same car. Thus, there are $20! - 5 \cdot 4 \cdot 3 \cdot 18!$ to seat the riders such that the certain two people are {\it not} in the same car.

\subsection*{28. A secretary works in a building located nine blocks east and eight blocks north of his home. Every day he walks $17$ blocks to work.}
\subsubsection*{(a) How many different routes are possible for him?}
Assuming the secretary may only walk one block in a cardinal direction at a time, we see that he must walk either east or north at any step; walking south or west would increase the total distance beyond $17$ blocks. Therefore, his route must consist of exactly nine block-steps eastward and exactly eight blocks northward, in any order. Then, there are $\frac{17!}{9! \cdot 8!}$ possible routes.

\subsubsection*{(b) How many different routes are possible if the one block in the easterly direction, which begins four blocks east and three blocks north of his home, is under water (and he can't swim)? ({\it Hint}: Count the routes that use the block under water).}
To use the block under water, the secretary must travel to $(4, 3)$, walk east to $(5, 3)$, then travel to $(9, 8)$. Using the same reasoning as before, we see there are $\frac{7!}{4! \cdot 3!}$ ways to make the first leg of this trip; for each, one way to make the second leg; for each, $\frac{10!}{5! \cdot 5!}$ ways to make the last leg. Then, there are $\frac{10! \cdot 7!}{5! \cdot 5! \cdot4! \cdot 3!}$ routes through the flooded block. Thus, there are $\frac{17!}{9! \cdot 8!} - \frac{10! \cdot 7!}{5! \cdot5! \cdot 4! \cdot 3!}$ routes which do {\it not} go through the flooded block.

\subsection*{32. Determine the number of $11$-permutations of the multiset $S = \{3 \cdot a, 4 \cdot b, 5 \cdot c\}$.}
The $11$-permutations can be formed by removing one of each element in turn from the multiset and permuting the results. Then, the number of $11$-permutations is $\frac{11!}{2! \cdot 4! \cdot 5!} + \frac{11!}{3! \cdot 3! \cdot 5!} + \frac{11!}{3! \cdot 4! \cdot 4!}$.

\subsection*{35. List all $3$-combinations and $4$-combinations of the multiset $S = \{2 \cdot a, 1 \cdot b, 3 \cdot c\}$.}
\begin{center}
  \begin{tabular}[c]{c || c}
    $3$-combinations & $4$-combinations\\
    \hline
    $\{0 \cdot a, 0 \cdot b, 3 \cdot c\}$ & $\{0 \cdot a, 1 \cdot b, 3 \cdot c\}$\\
    $\{0 \cdot a, 1 \cdot b, 2 \cdot c\}$ & $\{1 \cdot a, 0 \cdot b, 3 \cdot c\}$\\
    $\{1 \cdot a, 0 \cdot b, 2 \cdot c\}$ & $\{1 \cdot a, 1 \cdot b, 2 \cdot c\}$\\
    $\{1 \cdot a, 1 \cdot b, 1 \cdot c\}$ & $\{2 \cdot a, 0 \cdot b, 2 \cdot c\}$\\
    $\{2 \cdot a, 0 \cdot b, 1 \cdot c\}$ & $\{2 \cdot a, 1 \cdot b, 1 \cdot c\}$\\
    $\{2 \cdot a, 1 \cdot b, 0 \cdot c\}$ & \\
  \end{tabular}
\end{center}

\subsection*{39. There are $20$ identical sticks lined up in a row occupying $20$ distinct places. Six of them are to be chosen.}
\subsubsection*{(a) How many choices are there?}
$20 \choose 6$.

\subsubsection*{(b) How many choices are there if no two of the chosen sticks can be consecutive?}
TODO (answer is 5005)

\subsubsection*{(c) How many choices are there if there must be at least two sticks between each pair of chosen sticks?}
TODO (answer is 3003)

\subsection*{55. How many permutations are there of the letters in the words}
\subsubsection*{(a) TRISKAIDEKAPHOBIA?}
TRISKAIDEKAPHOBIA has $17$ letters and corresponds to the multiset
\begin{align*}
\{3 \cdot A, 1 \cdot B, 1 \cdot D, 1 \cdot E, 1 \cdot H, 3 \cdot I, 2 \cdot K, 1 \cdot O, 1 \cdot P, 1 \cdot R, 1 \cdot S, 1 \cdot T\}.
\end{align*}

It thus has $\frac{17!}{3! \cdot 1! \cdot 1! \cdot 1! \cdot 1! \cdot 3! \cdot 2! \cdot 1! \cdot 1! \cdot 1! \cdot 1! \cdot 1!} = \frac{17!}{3! \cdot 3! \cdot 2!}$ letter permutations.

\end{document}
