\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage[md]{titlesec}

\renewcommand\thesection{{\bf Part \Roman{section}:}}
\renewcommand\thesubsection{{\bf\arabic{subsection}.}}
\renewcommand\thesubsubsection{\alph{subsubsection}.}

\title{MATH 4350 -- Midterm}
\author{Stuart Olsen}
\date{March 23, 2015}

\begin{document}
\maketitle


\section{Answer all of the following short questions.}
\subsection{Answer the following for the set $S = \{1, 2, 3, 4, 5\}$.}
\subsubsection{Write two $3$-permutations of $S$ that are the same $3$-combination.}
\begin{itemize}
\item $(1, 2, 3)$
\item $(1, 3, 2)$
\end{itemize}

\subsubsection{Write two $3$-permutations of $S$ that are the same cyclic $3$-permutation.}
\begin{itemize}
\item $(1, 2, 3)$
\item $(3, 1, 2)$
\end{itemize}

\subsubsection{Write a $5$-permutations of $S$ that is a derangement of $S$.}
\begin{itemize}
\item $(5, 1, 2, 3, 4)$
\end{itemize}

\subsubsection{Using the known formulas, count the total number of $3$-permutations, $3$-combinations, cyclic $3$-permutations, and derangements of $S$. Label each clearly.}
\begin{align*}
  (5)_3 &= \frac{5!}{(5-3)!} = 5 \cdot 4 \cdot 3 &= 60 \tag{$3$-permutations}\\
  {5 \choose 3} &= \frac{5!}{3!\,(5-3)!} = \frac{5 \cdot 4 \cdot 3}{3 \cdot 2 \cdot 1} &= 10 \tag{$3$-combinations}\\[0.7em]
  \frac{(5)_3}{3} &= \frac{5!}{3(5-3)!} = \frac{5 \cdot 4 \cdot 3}{3} &= 20 \tag{cyclic $3$-permutations}\\[0.6em]
  D_5 &= 5!(1 - \frac{1}{1!} + \frac{1}{2!} - \frac{1}{3!} + \frac{1}{4!} - \frac{1}{5!}) &= 44 \tag{derangements}
\end{align*}


\subsection{Answer the following for the multiset $T = \{3 \cdot a, 3 \cdot b, 4 \cdot c\}$.}
\subsubsection{How many permutations of $T$ are there?}
\begin{align*}
  \frac{(3+3+4)!}{3!\,3!\,4!} &= \frac{10!}{36 \cdot 4!} = 4200.
\end{align*}

\subsubsection{How many $3$-combinations of $T$ are there?}
Because $3$ is less than or equal to the count of each type in $T$, we can use the formula
\begin{align*}
  {{3 + 3 - 1} \choose {3}} &= {{5} \choose {3}} = 10.
\end{align*}


\subsection{How many people are required to guarantee that two are born on the same day?}
There are $7$ days in each week, $31$ days in the longest month, and $366$ days in the longest year. Then, by the pigeonhole principle, $8$ people are required to guarantee that two are born on the same day of the week, $32$ people are required to guarantee that two are born on the same day of the month, and $367$ people are required to guarantee that two are born on the same day of the year.


\subsection{A group of $4$ men and $6$ women is to form a committee consisting of $5$ members.}
\subsubsection{How many committees can be formed?}
\begin{align*}
  {{4 + 6} \choose {5}} &= {{10} \choose {5}} = 252.
\end{align*}

\subsubsection{How many committees can be formed if there must be at least $2$ men or at least $2$ women?}
By excluding the choices giving zero or one men and those giving zero or one women (which are mutually exclusive), we have
\begin{align*}
  {{4 + 6} \choose {5}} - \left[ {{4} \choose {0}} \cdot {{6} \choose {5}}
    + {{4} \choose {1}} \cdot {{6} \choose {4}}
    + {{6} \choose {0}} \cdot {{4} \choose {5}}
    + {{6} \choose {1}} \cdot {{4} \choose {4}} \right]
  &= 252 - \left[ 1 \cdot 6 + 4 \cdot 15 + 1 \cdot 0 + 6 \cdot 1 \right]\\
  &= 252 - \left[ 6 + 60 + 0 + 6 \right]\\
  &= 252 - 72\\
  &= 180.
\end{align*}


\subsection{}
\subsubsection{Find the coefficient of the $x^6$ term of $(x-2)^7$ using the binomial theorem.}
\begin{align*}
  {7 \choose 1} \cdot (-2)^1 &= -14.
\end{align*}

\subsubsection{Find the coefficient of the $x^4y^2$ term of $(x+y-2)^7$ using the multinomial theorem.}
\begin{align*}
  {7 \choose {4\ 2\ 1}} \cdot (-2)^1 &= -210.
\end{align*}


\subsection{How many integers between $1$ and $200$ are not divisible by $3$, $5$, and $7$?}
The integers which {\em are} divisible by $3$, $5$, and $7$ are exactly those divisible by $\mathrm{lcm}(3, 5, 7) = 105$. There is one such integer between $1$ and $200$: $105$. Then, the number of integers between $1$ and $200$ {\em not} divisible by $3$, $5$, and $7$ is $|(1, 200)| - 1 = 198 - 1 = 197$.


\subsection{Prove that for all real numbers $r$ and integers $k$ the following relationship holds:}
\begin{align*}
  {{-r} \choose {k}} &= (-1)^k {{r + k - 1} \choose {k}}
\end{align*}

For $r \in \mathbb{R}$ and $k \in \mathbb{Z}$, we have
\begin{align*}
  {r \choose k} = \begin{dcases*}
    \frac{r(r-1)\cdots(r-k+1)}{k!} & if $k \ge 1$\\
    1 & if $k = 0$\\
    0 & if $k \le -1$.\\
  \end{dcases*}
\end{align*}

We proceed by case analysis on $k$:
\begin{itemize}
\item $k \le -1$:
  \begin{align*}
    {{-r} \choose {k}} &= 0\\
    &= (-1)^k \cdot 0\\
    &= (-1)^k {{r + k - 1} \choose {k}}
  \end{align*}
\item $k = 0$:
  \begin{align*}
    {{-r} \choose {k}} &= 1\\
    &= 1 \cdot 1\\
    &= (-1)^k \cdot 1\\
    &= (-1)^k {{r + k - 1} \choose {k}}\\
  \end{align*}
\item $k \ge 1$:
  \begin{align*}
    {{-r} \choose {k}} &= \frac{-r(-r-1)\cdots(-r-k+1)}{k!}\\
    &= \frac{(-r-k+1)(-r-k+2)\cdots{-r}}{k!}\\
    &= (-1)^k \frac{(r+k-1)(r+k-2)\cdots{r}}{k!}\\
    &= (-1)^k \frac{(r+k-1)((r+k-1)-1)\cdots((r+k-1)-k+1)}{k!}\\
    &= (-1)^k {{r + k - 1} \choose {k}}
  \end{align*}
\end{itemize}



\section{Answer any {\em three} of the following questions.}
% \subsection{}
% \subsubsection{State the binomial theorem.}
% \subsubsection{Provide an algebraic proof that $\sum\limits_{k=0}^n{\frac{{n \choose k}}{k + 1}} = \frac{2^{n+1} - 1}{n + 1}$.}
\setcounter{subsection}{1}
\subsection{Answer the following for the grid below.}
\subsubsection{How many shortest paths are there from $A$ to $B$?}
Each shortest path will consist of $4$ steps north and $8$ steps east. Because the paths differ only in the order in which the steps are taken and two steps in a given direction are indistinguishable, the paths are best modeled as permutations of the multiset $\{4 \cdot N, 8 \cdot E\}$. The number of shortest paths is then
\begin{align*}
  \frac{(4+8)!}{4!\,8!} &= \frac{12!}{4!\,8!} = 495.
\end{align*}

\subsubsection{How many shortest paths are there from $A$ to $C$ to $B$?}
By the same construction as above, there are $\frac{(1+4)!}{1!\,4!} = 5$ shortest paths from $A$ to $C$ and $\frac{(3+4)!}{3!\,4!} = 35$ shortest paths from $C$ to $B$.

Because each shortest path from $A$ to $C$ to $B$ is comprised of a shortest path from $A$ to $C$ and a shortest path from $C$ to $B$, there are $5 \cdot 35 = 175$ shortest paths from $A$ to $C$ to $B$.

\subsubsection{How many shortest paths are there from $A$ to $B$ not passing through $C$ or $D$?}
By the same construction as above, there are $\frac{(2+6)!}{2!\,6!} \cdot \frac{(2+2)!}{2!\,2!} = 168$ shortest paths from $A$ to $D$ to $B$ and $\frac{(1+4)!}{1!\,4!} \cdot \frac{(1+2)!}{1!\,2!} \cdot \frac{(2+2)!}{2!\,2!} = 90$ shortest paths from $A$ to $C$ to $D$ to $B$.

By the inclusion-exclusion principle, the number of shortest paths from $A$ to $B$ which do not pass through $C$ or $D$ is then $495 - (175 + 168 - 90) = 242$.

\end{document}
