\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

\title{MATH 4350 -- Problem Set 1}
\author{Stuart Olsen}

\begin{document}
\maketitle

\section*{Chapter 2.7}
\subsection*{16. Prove that ${n \choose r} = {n \choose {n - r}}$ by using a combinatorial argument and not the values of these numbers as given in Theorem 2.3.1.}
$n \choose r$ denotes the number of ways to choose $r$ elements of $n$; that is, to partition $n$ elements into a subset of $r$ ``chosen'' elements and a subset of $n - r$ ``not chosen'' elements. By exchanging the labels ``chosen'' and ``not chosen,'' we see that the number of ways to choose $n - r$ elements of $n$ is the same as the number of ways to choose $r$ elements, i.e. ${n \choose r} = {n \choose {n - r}}$.

\subsection*{29. Let $S$ be a multiset with repetition numbers $n_1, n_2, \ldots, n_k$, where $n_1 = 1$. Let $n = n_2 + \cdots + n_k$. Prove that the number of circular permutations of $S$ equals $\frac{n!}{n_2! \cdots n_k!}$.}
Note that $\left| S \right| = n + n_1$. By Theorem 2.4.2, there are $\frac{(n+n_1)!}{n_1!n_2! \cdots n_k!}$ non-circular permutations of $S$. By Theorem 2.2.2, then, there are $\frac{(n+n_1)!}{(n+n_1) \cdot n_1!n_2! \cdots n_k!} = \frac{(n+n_1 - 1)!}{n_1!n_2! \cdots n_k!} = \frac{n!}{n_2! \cdots n_k!}$ circular permutations of $S$.

\subsection*{48. Prove that the number of permutations of $m$ As and at most $n$ Bs equals ${m + n + 1} \choose {m + 1}$.}
By Theorem 2.4.2, the number of permutations of $m$ As and exactly $n$ Bs is given by $\frac{(m+n)!}{m!n!}$. Then, the number of permutations of $m$ As and {\it at most} $n$ Bs is $\sum_{i=0}^n \frac{(m+i)!}{m!i!}$. Letting $f(n)$ denote this value, we proceed to prove $f(n) = {{m + n + 1} \choose {m + 1}}$ by induction:

\begin{itemize}
\item $k = 0$:

Evaluating $f(k)$ and ${{m+k+1} \choose {m+1}}$ for $k = 0$ yields:

  \begin{align*}
    f(k) &= f(0)\\
    &= \sum_{i=0}^0 \frac{(m+0)!}{m!0!} & {{m+k+1} \choose {m+1}} &= {{m+0+1} \choose {m+1}}\\
    &= \frac{m!}{m!} & &= {{m+1} \choose {m+1}}\\
    &= 1 & &= 1.\\
  \end{align*}

Then, $f(k) = {{m+k+1} \choose {m+1}}$.

\item $k > 1$:

Take as the inductive hypothesis that $f(k-1) = {{m+(k-1)+1} \choose {m+1}}$. Then,

\begin{align*}
  f(k-1) &= {{m+(k-1)+1} \choose {m+1}}\\
  \sum_{i=0}^{k-1} \frac{(m+i)!}{m!i!} &= \frac{(m+k)!}{(m+1)!(k-1)!}\\
  \sum_{i=0}^{k-1} \frac{(m+i)!}{m!i!} + \frac{(m+k)!}{m!k!} &= \frac{(m+k)!}{(m+1)!(k-1)!} + \frac{(m+k)!}{m!k!}\\
  \sum_{i=0}^{k} \frac{(m+i)!}{m!i!} &= \frac{(m+k)!}{m!k!} \cdot \frac{k}{m+1} + \frac{(m+k)!}{m!k!}\\
  f(k) &= \frac{(m+k)!}{m!k!} \cdot \left( \frac{k}{m+1} + 1 \right)\\
  f(k) &= \frac{(m+k)!}{m!k!} \cdot \frac{k+(m+1)}{m+1}\\
  f(k) &= \frac{(m+k)!}{m!k!} \cdot \frac{(m+k)+1}{m+1}\\
  f(k) &= \frac{(m+k+1)!}{(m+1)!k!}\\
  f(k) &= {{m+k+1} \choose {m+1}}.
\end{align*}
\end{itemize}

Therefore, by induction, the number of permutations of $m$ As and at most $n$ Bs equals ${m + n + 1} \choose {m + 1}$. $\blacksquare$

\subsection*{53. Find a one-to-one correspondance between the permutations of the set $\{1, 2, \ldots, n\}$ and the towers $A_0 \subset A_1 \subset A_2 \subset \cdots \subset A_n$ where $|A_k| = k$ for $k = 0, 1, 2, \ldots, n$.}
TODO (``tower?'')

\section*{Provide an algebraic proof of Pascal's Formula, ${n \choose k} = {{n - 1} \choose k} + {{n - 1} \choose {k - 1}}$.}

\begin{align*}
  {n \choose k} &= \frac{n!}{k!(n-k)!}\\
                &= \frac{(n-1)! \cdot n}{k!(n-k)!}\\
                &= \frac{(n-1)!}{k!(n-k)!} \cdot (n - k + k)\\
                &= \frac{(n-1)! \cdot (n-k)}{k!(n-k)!} + \frac{(n-1)! \cdot k}{k!(n-k)!}\\
                &= \frac{(n-1)!}{k!(n-1-k)!} + \frac{(n-1)!}{(k-1)!(n-k)!}\\
                &= {{n-1} \choose k} + {{n-1} \choose {k-1}}\\ \blacksquare
\end{align*}

\section*{Provide a combinatorial proof of Gauss' Formula, $1 + 2 + 3 + \cdots + (n - 1) = {n \choose 2}$.}
Consider a set $S$ with $n$ elements $x_1$ to $x_n$. The number of unordered pairs containing $x_1$ is $n-1$: one for each of $x_2$ through $x_n$. The number of pairs containing $x_2$ (discounting those previously counted) is $n-2$: one for each of $x_3$ through $x_n$. Continuing this way, we count one new pair when considering $x_{n-1}$, and no new pairs when considering $x_n$. Thus, the total number of unordered pairs which can be formed from $S$, i.e. $n \choose 2$, is $(n-1) + (n-2) + \cdots + 1 + 0$, or $1 + 2 + 3 + \cdots + (n-1)$.

\section*{Explain the proof of Theorem 2.4.2 (in detail, in your own words).}
The proof works by allocating $n$ indices for the $n$-multiset's elements one ``type'' at a time. For each type $a_k$ in turn, with count $n_k$, the proof {\it chooses} $n_k$ of the remaining indices and populates them with the $a_k$ in no particular order. The proof uses the multiplication principle to accumulate the numbers of these choices, and arithmetically demonstrates the result to be equivalent to the given formula.

\end{document}
