\documentclass{homework}
\usepackage{fullpage}
\usepackage{amsthm}

\usepackage{environ}
\NewEnviron{caselist}{%
  \begin{itemize}%
    \BODY%
  \end{itemize}%
}
\newcommand{\case}[2]{%
  \item {#1}:

    {#2}%
}

\newcommand{\incomplete}[1]{%
  {\it Incomplete.}
}



\title{MATH 4350 -- Problem Set 2}
\author{Stuart Olsen}

\begin{document}
\maketitle


\section*{Chapter 3.4}
\problem{2}{
  Concerning Application 5, show that if $100$ integers are chosen from $\{1, 2, \ldots, 200\}$ and one of the integers chosen is less than $16$, then there are two chosen numbers such that one of them is divisible by the other.
}
\begin{proof}
  Let $S$ be a set of $100$ integers chosen from $\{1, 2, \ldots, 200\}$ and indexed by some set $J$. Further, suppose $S$ has some element $x$ less than $16$. Each element of $S$ may be written in the form $2^k \times a$, where $k$ is a nonnegative integer and $a$ is odd; let $k_i$ and $a_i$ denote this $k$ and $a$ for each $s_i$ in $S$, respectively.

  Consider whether there exist distinct $i, j \in J$ such that $a_i = a_j$:
  \begin{caselist}
    \case{yes ($\qsome{i,j\in J}{a_i=a_j}$)}{
      Consider whether $k_i \ge k_j$:
      \begin{caselist}
        \case{yes ($k_i \ge k_j$)}{
          Because $k_i \ge k_j$, there is some nonnegative integer $d$ such that $k_i = k_j + d$. Then,
          %%% FIXME: Center
          \begin{align*}
            s_i &= 2^{k_i} \times a_i\\
                &= 2^{k_j + d} \times a_i\\
                &= 2^{k_j + d} \times a_j\\
                &= 2^d (2^{k_j} \times a_j)\\
                &= 2^d s_j.
          \end{align*}
          Because $d$ is a nonnegative integer, so is $2^d$; then, $s_j \divides s_i$.
        }
        \case{no ($k_i \ngeq k_j$)}{
          Because $k_i \ngeq k_j$, $k_j \ge k_i$; by congruent reasoning to the above, $s_i \divides s_j$.
        }
      \end{caselist}
    }

    \case{no ($\nqsome{i,j\in J}{a_i=a_j}$)}{
      There are no two $i, j \in J$ such that $a_i = a_j$, so there are $100$ distinct $a_i$. Then, because there are $100$ odd numbers in $\{1, 2, \ldots, 200\}$, each such odd number appears as exactly one $a_i$.

      Consider the following exhaustive cases of $x$:
      \begin{caselist}
        \case{$x \in \{1\}$}{
          $x$ divides every other element of $S$; pick one.
        }
        \case{$x \in \{3, 5, 7, 9, 11, 13, 15\}$}{
          $13x$ is odd, greater than $15$, and less than $200$; therefore, there is some $i$ such that $s_i = 2^{k_i} \times 13x$, and thus that $x \divides s_i$.
        }
        \case{$x \in \{2, 4, 8\}$}{
          There are some $i, j, l, m \in J$ such that $a_i = 81$, $a_j = 27$, $a_l = 9$, and $a_m = 3$. If $k_j < 1$, then $s_j \divides s_i$; otherwise, if $k_l < 2$, then $s_l \divides s_j$; otherwise, if $k_m < 3$, then $s_m \divides s_l$; otherwise, $k_m$ is at least $3$, so $8 \divides s_m$, and so $x \divides s_m$.
        }
        \case{$x \in \{6, 10, 14\}$}{
          Let $i$ be the element of $J$ such that $x = s_i$. Note that $3a_i$ is an odd number between $1$ and $200$. Consider whether $3a_i \in S$:
          \begin{caselist}
            \case{$3a_i \in S$}{
              Let $j$ be the element of $J$ such that $3a_i = s_j$. Note that $3s_j$ is an odd number between $1$ and $200$. Then, there is some $l \in J$ such that $a_l = 3s_j$, and thus that $s_j \divides a_l$, and thus that $s_j \divides s_l$.
            }
            \case{$3a_i \notin S$}{
              Let $j$ be the element of $J$ such that $3a_i = a_j$. Then, because $a_j \notin S$, $k_j > 0$, and so $2a_j \divides s_j$, and so $6a_i \divides s_j$. Because each of $\{6, 10, 14\}$ is of the form $2^1 \times a$, we have $3s_i \divides s_j$, and thus $s_i \divides s_j$.
            }
          \end{caselist}
        }
        \case{$x \in \{12\}$}{
          If $9 \in S$, we may defer without loss of generality to the above case where $x \in \{3, 5, 7, 9, 11, 13, 15\}$. Otherwise, consider whether $18 \in S$:
          \begin{caselist}
            \case{$18 \in S$}{
              If $27 \in S$, then $27$ divides the term $s_i$ such that $a_i = 81$. Otherwise, a multiple of $54$ must be in $S$; in this case, $18$ divides that number.
            }
            \case{$18 \notin S$}{
              Because $9 \notin S$ and $18 \notin S$, some multiple of $2^2 \times 9 = 36$ must be in $S$; in this case, $x$ divides that number.
            }
          \end{caselist}
        }
      \end{caselist}
    }
  \end{caselist}
\end{proof}


\problem{8}{
  Use the pigeonhole principle to prove that the decimal expansion of a rational number $m/n$ eventually is repeating. For example,
  \begin{align*}
    \frac{34478}{99900} = 0.345125125125\cdots.
  \end{align*}
}
\incomplete{
  We may assume without loss of generality that $n$ is a positive integer and $m$ is a nonnegative integer less than $n$.
}


\problem{15}{
  Prove that, for any $n+1$ integers $a_1, a_2, \ldots, a_{n+1}$, there exist two of the integers $a_i$ and $a_j$ with $i \ne j$ such that $a_i - a_j$ is divisible by $n$.
}
\incomplete{
}


\end{document}
